Insertion Sort Analysis in Java

### 2015-02-03 Programming Project A/Insertion Sort Analysis in Java.tex
added content to paper

%% BEGIN PROCEDURE
\section{Procedure}
An insertion sort can be implemented in a multitude of languages using the pseudocode provided below.
\newline
\textbf{Insertion Sort Pre-Condition}: A is a non-empty array of data with a natural order.
\newline
\textbf{Insertion Sort Post-Condition}: A' is a permutation of A (containing all the same elements) in strictly non-decreasing order.
\begin{algorithm}
\caption {\textsc{Insertion-Sort}(A)}
\label{algo:insertionsort}
\begin{algorithmic}[1]
\Procedure{Insertion-Sort}{A}
\State{$i = 2$}
\While{$i$ upto $A.length$}
	\State{$key = A[i]$}
	\State{$j = i - 1$}
	\While{$j$ downto $1$ and $key < A[j]$}
		\State{$A[j + 1] = A[j]$}
		\State{$j = j - 1$}
	\EndWhile
	\State{$A[j+1] = key$}
	\State{$i = i + 1$}
\EndWhile
\EndProcedure
\end{algorithmic}
\end{algorithm}
\newline
\textbf{Outer-Loop Invariant}: The subarray A'[1 ... i - 1] contains all the same elements as the subarray A[1 .. i - 1].
\newline 
\textbf{Outer-Loop Initialization}: The outer-loop invariant holds because A'[1 ... i - 1] and A[1 ... i - 1] both contain the same one element.
\newline
\textbf{Outer-Loop Maintenance}: The outer-loop invariant holds because A'[1 ... i - 1] and A[1 ... i - 1] both contain the same elements, although they maybe in different orders.
\newline
\textbf{Outer-Loop Termination}: When the outer-loop terminates, i = A.length, which implies that the entire array has been traversed and the guard has been negated. The negation of the guard implies that A'[1 ... i - 1] contains all the elements in A[1 ... i - 1].
\newline
\newline
\textbf{Inner-Loop Invariant}: A'[1 ... j] is sorted in strictly non-decreasing order.
\newline
\textbf{Inner-Loop Initialization}: Before the first iteration of the loop, j = 1, meaning the subarray A'[1 ... j] contains exactly one element, which is already sorted.
\newline
\textbf{Inner-Loop Maintenance}: At the beginning of each iteration of the loop the inner-loop invariants holds because j counts down from i and A'[j+1] is swapped with A'[j] only if A'[j+1] is less than A[j].
\newline
\textbf{Inner-Loop Termination}: The negation of the implies that j = A.length and A'[1 ... j] has been entirely traversed and sorted in strictly non-decreasing order which maintains the inner-loop invariant. 
\newline
\newline
\textbf{Conclusion}: The termination of both the inner and outer loops implies that the entire array has been traversed, A' is a permutation of A containing all the same elements in strictly non-decreasing order. This satisfies the post condition.
%% END PROCEDURE

@@

%% BEGIN EXPERIMENTAL ANALYSIS
\section{Experimental Analysis}
 The insertion sort demonstrated in Algorithm 1 was implemented in Java and executed on an HP SprectreXT TouchSmart with 4 Core Intel i7 processor clocked at 1.9GHz running Ubuntu Gnome 14.10 64-bit.
 \begin{figure}[!]
\begin{center}
\includegraphics[width=0.7\linewidth]{insertion_sort_time_analysis.pdf}
\end{center}
\captionof{figure}{Insertion Sort Time Analysis}
\label{fig:insertionsorttimeanalysis}
\end{figure}

%% END EXPERIMENTAL ANALYSIS

@@

%% BEGIN APPENDIX
\section*{Appendix}
\lstinputlisting[caption=TestDriver,
label={TestDriver.java},
breaklines=true,
]{Code/TestDriver.java}
\lstinputlisting[caption=Debug,
label={Debug.java},
breaklines=true,
]{Code/Debug.java}
\lstinputlisting[caption=DummyData,
label={DummyData.java},
breaklines=true,
]{Code/DummyData.java}
\lstinputlisting[caption=Stopwatch,
label={Stopwatch.java}
breaklines=true,
]{Code/Stopwatch.java}
\lstinputlisting[caption=Sort,
label={Sort.java},
breaklines=true,
]{Code/Sort.java}
%% END APPENDIX

>>>>

%% BEGIN PROCEDURE
\section{Procedure}
An insertion sort can be implemented in a multitude of languages using the pseudocode provided below.
\newline
\textbf{Insertion Sort Pre-Condition}: A is a non-empty array of data with a natural order.
\newline
\textbf{Insertion Sort Post-Condition}: A' is a permutation of A (containing all the same elements) in strictly non-decreasing order.
\begin{algorithm}
\caption {\textsc{Insertion-Sort}(A)}
\label{algo:insertionsort}
\begin{algorithmic}[1]
\Procedure{Insertion-Sort}{A}
\If{$A.length <= 1$}
\State{\Return{$A$}}
\EndIf
\State{$i = 2$}
\While{$i$ upto $A.length$}
	\State{$key = A[i]$}
	\State{$j = i - 1$}
	\While{$j$ downto $1$ and $key < A[j]$}
		\State{$A[j + 1] = A[j]$}
		\State{$j = j - 1$}
	\EndWhile
	\State{$A[j+1] = key$}
	\State{$i = i + 1$}
\EndWhile
\EndProcedure
\end{algorithmic}
\end{algorithm}
\newline
\textbf{Outer-Loop Invariant}: The subarray A'[1 ... i - 1] contains all the same elements as the subarray A[1 .. i - 1].
\newline 
\textbf{Outer-Loop Initialization}: The outer-loop invariant holds because A'[1 ... i - 1] and A[1 ... i - 1] both contain the same one element.
\newline
\textbf{Outer-Loop Maintenance}: The outer-loop invariant holds because A'[1 ... i - 1] and A[1 ... i - 1] both contain the same elements, although they maybe in different orders.
\newline
\textbf{Outer-Loop Termination}: When the outer-loop terminates, i = A.length, which implies that the entire array has been traversed and the guard has been negated. The negation of the guard implies that A'[1 ... i - 1] contains all the elements in A[1 ... i - 1].
\newline
\newline
\textbf{Inner-Loop Invariant}: A'[1 ... j] is sorted in strictly non-decreasing order.
\newline
\textbf{Inner-Loop Initialization}: Before the first iteration of the loop, j = 1, meaning the subarray A'[1 ... j] contains exactly one element, which is already sorted.
\newline
\textbf{Inner-Loop Maintenance}: At the beginning of each iteration of the loop the inner-loop invariants holds because j counts down from i and A'[j+1] is swapped with A'[j] only if A'[j+1] is less than A[j].
\newline
\textbf{Inner-Loop Termination}: The negation of the implies that j = A.length and A'[1 ... j] has been entirely traversed and sorted in strictly non-decreasing order which maintains the inner-loop invariant. 
\newline
\newline
\textbf{Conclusion}: The termination of both the inner and outer loops implies that the entire array has been traversed, A' is a permutation of A containing all the same elements in strictly non-decreasing order. This satisfies the post condition.
%% END PROCEDURE

@@

%% BEGIN EXPERIMENTAL ANALYSIS
\section{Experimental Analysis}
 The insertion sort demonstrated in Algorithm 1 was implemented in Java and executed on an HP SprectreXT TouchSmart with 4 Core Intel i7 processor clocked at 1.9GHz running Ubuntu Gnome 14.10 64-bit.
 \begin{figure}[!]
\begin{center}
\includegraphics[scale=.75]{insertion_sort_time_analysis.pdf}
\end{center}
\captionof{figure}{Insertion Sort Time Analysis}
\label{fig:insertionsorttimeanalysis}
\end{figure}

The expected growth of insertion sort as the number of elements (n) grows large can be represented as $ \theta(n^2) $ where $ \theta() $ represents the asymptotically tightly bound running time. At $n_0$ the algorithm took 0.0003 milliseconds to complete. At $n_1$ the algorithm also took 0.0003 milliseconds to complete. For both these values of $n$, the algorithm runs at a constant time because no for-loops are executed. As $n$ grows larger the time to complete the experimental data correlates quite accurately with the expected growth. For $n_{1*10^6} $ the data matches up perfectly. This is to be expected however, because as n grows larger the constants and lower orders of the actual running time of the insertion sort start to effect the the running time less and less because the highest order of $n^2$ is so large. 
%% END EXPERIMENTAL ANALYSIS

@@

%% BEGIN APPENDIX
\section*{Appendix}
\lstinputlisting[caption=TestDriver,
label={TestDriver.java},
breaklines=true,
]{Code/TestDriver.java}
\lstinputlisting[caption=Debug,
label={Debug.java},
breaklines=true,
]{Code/Debug.java}
\lstinputlisting[caption=DummyData,
label={DummyData.java},
breaklines=true,
]{Code/DummyData.java}

The class Stopwatch has not been altered from its original form.
\newline
\lstinputlisting[caption=Stopwatch,
label={Stopwatch.java}
breaklines=true,
]{Code/Stopwatch.java}
\lstinputlisting[caption=Sort,
label={Sort.java},
breaklines=true,
]{Code/Sort.java}
%% END APPENDIX
